---
title: "Covariate Selection with the Horseshoe Prior"
package: jmpost
bibliography: '`r system.file("REFERENCES.bib", package = "jmpost")`'
csl: '`r system.file("jss.csl", package = "jmpost")`'
output:
  rmarkdown::html_vignette:
          toc: true
link-citations: true
vignette: >
  %\VignetteIndexEntry{Covariate Selection with the Horseshoe Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(jmpost)
```

## Overview

Covariate selection in joint models is often difficult because each candidate
covariate can interact with the survival model, the longitudinal model, and the
association between them. The regularized horseshoe prior provides a Bayesian
shrinkage approach: include the candidate covariates in one model, shrink most
survival coefficients strongly towards zero, and allow a smaller number of
coefficients to escape that shrinkage when supported by the data.

In `jmpost`, this workflow is available for the survival covariate coefficients
through `prior_horseshoe()`. This vignette starts from the prior specification
and then shows a complete joint model example in which the survival covariates
use the horseshoe prior. The fitted model can be inspected through the posterior
coefficient draws and through the shrinkage factors extracted with `shrinkage()`.

## Theory

The horseshoe prior is a global-local shrinkage prior [@CarvalhoEtAl2010]. A
global parameter shrinks all coefficients towards zero, while coefficient-level
local parameters allow selected coefficients to remain large. The regularized
horseshoe [@PiironenVehtari2017] adds a slab component that regularizes
coefficients that escape the main shrinkage. This is the parameterization used
by `jmpost`, following the same notation as `brms` [@Buerkner2017].
Details are provided in the statistical specifications vignette [here](statistical-specification.html).

The hyperparameters map to `prior_horseshoe()`
arguments as follows:

- `df`: local degrees of freedom, $\nu_{\lambda}$
- `df_global`: global degrees of freedom, $\nu_{\tau}$
- `df_slab`: slab degrees of freedom, $\nu_c$
- `scale_global`: global scale, $s_{\tau}$
- `scale_slab`: slab scale, $s_c$

The local shrinkage parameters $\lambda_j$ let individual coefficients escape
the global shrinkage $\tau$. The slab scale $s_c$ and slab degrees of freedom
$\nu_c$ control how much very large coefficients are regularized. Smaller
$s_{\tau}$ values encode stronger prior belief that only a small number of
candidate covariates are relevant.

A useful posterior diagnostic is the shrinkage factor

$$
\kappa_j =
\frac{1}{1 + \tau^2\lambda_j^2 / c^2}.
$$

Values of $\kappa_j$ close to 1 indicate strong shrinkage towards zero. Values
close to 0 indicate little shrinkage. The shrinkage factor should be read
together with the coefficient posterior: it is evidence about how strongly the
prior-likelihood combination pulled a coefficient towards zero, not a
standalone posterior inclusion probability.

## Simulate Example Data

We first simulate data from a joint model with a random-slope longitudinal
submodel and a Weibull proportional hazards survival submodel. The survival
model contains one categorical and one continuous candidate covariate. The
categorical covariate contributes two columns to the design matrix, so the
survival model has three covariate coefficients in total.

```{r simulate-data}
set.seed(129)
sim_data <- SimJointData(
    design = list(
        SimGroup(50, "Arm-A", "Study-X"),
        SimGroup(50, "Arm-B", "Study-X")
    ),
    longitudinal = SimLongitudinalRandomSlope(
        times = c(1, 20, 50, 100, 150, 200, 250, 300),
        intercept = 30,
        slope_mu = c(1, 2),
        slope_sigma = 0.2,
        sigma = 20,
        link_dsld = 0.1
    ),
    survival = SimSurvivalWeibullPH(
        lambda = 1 / 300,
        gamma = 0.97,
        time_max = 2000,
        time_step = 1,
        lambda_cen = 1 / 9000,
        beta_cat = c(
            "A" = 0,
            "B" = 0.3,
            "C" = 0.7
        ),
        beta_cont = 0.3
    )
)
```

Then we prepare the corresponding objects:

```{r prepare-data}
os_data <- sim_data@survival
long_data <- sim_data@longitudinal

joint_data <- DataJoint(
    subject = DataSubject(
        data = os_data,
        subject = "subject",
        arm = "arm",
        study = "study"
    ),
    survival = DataSurvival(
        data = os_data,
        formula = Surv(time, event) ~ cov_cat + cov_cont
    ),
    longitudinal = DataLongitudinal(
        data = long_data,
        formula = sld ~ time,
        threshold = 5
    )
)
```

Before setting priors or interpreting the output, it is a good habit to check the survival design
matrix. With `covariates()` we can access the column names, which map to the
corresponding survival coefficients that receive the horseshoe prior.

```{r design-matrix}
head(model.matrix(joint_data@survival))
survival_covariates <- covariates(joint_data@survival)
survival_covariates
```

## Fit a Joint Model

Now we define a joint model. The longitudinal model and the association link use
standard priors. The survival model uses `prior_horseshoe()` for the vector of
survival covariate coefficients:

```{r model-specification}
joint_model <- JointModel(
    longitudinal = LongitudinalRandomSlope(),
    survival = SurvivalWeibullPH(
        beta = prior_horseshoe(
            df = 1,
            df_global = 1,
            df_slab = 4,
            scale_global = 0.3,
            scale_slab = 2
        )
    ),
    link = linkDSLD()
)
```

The choice above uses half-Cauchy priors for the local and global shrinkage
parameters, because `df = 1` and `df_global = 1`. The relatively small
`scale_global = 0.3` favours sparse survival effects in this small example,
while `scale_slab = 2` still allows meaningfully large log-hazard coefficients
when the data support them.

The following code fits the model. In a serious analysis, increase the number
of warmup and sampling iterations and check convergence carefully.

```{r fit-model}
fit <- sampleStanModel(
    joint_model,
    data = joint_data,
    iter_warmup = 500,
    iter_sampling = 500,
    chains = 4,
    parallel_chains = 4,
    seed = 325,
    refresh = 0,
    show_exceptions = FALSE,
    show_messages = FALSE
)
```

## Inspect Coefficients

After fitting, let's inspect the coefficient posterior and standard MCMC diagnostics.
The survival covariate coefficients are stored in `beta_os_cov`.

```{r inspect-coefficients}
stan_fit <- cmdstanr::as.CmdStanMCMC(fit)

stan_fit$summary(
    variables = c(
        "beta_os_cov",
        "prior_global_beta_os_cov",
        "prior_slab_beta_os_cov"
    )
)
```

The coefficient names come from the design matrix:

```{r name-coefficients}
beta_summary <- stan_fit$summary("beta_os_cov")
beta_summary$covariate <- survival_covariates
beta_summary[, c("covariate", "median", "q5", "q95", "rhat", "ess_bulk")]
```

Coefficients whose posterior remains close to zero and whose shrinkage factors
are close to 1 are natural candidates to treat as weakly supported covariates.
Coefficients whose posterior is away from zero and whose shrinkage factors are
closer to 0 have escaped shrinkage and are more strongly supported by the model.

## Extract and Plot Shrinkage Factors

Let's use the `shrinkage()` function to extract the posterior draws of $\kappa_j$. The returned
draws are named with the survival covariate names, so they can be plotted or
summarised directly.

```{r shrinkage-factors}
shrinkage_draws <- shrinkage(fit)
posterior::variables(shrinkage_draws)

summary(shrinkage_draws)
```

One compact visual summary is a density plot of the shrinkage factors.

```{r shrinkage-density, fig.width=7, fig.height=4}
library(bayesplot)

mcmc_dens_overlay(shrinkage_draws) +
    ggplot2::labs(
        x = "Shrinkage factor",
        y = "Posterior density"
    )
```

We see that both coefficients for the categorical covariate have shrinkage factors near 1, while the continuous 
covariate's shrinkage factor is a bit more distributed towards smaller values.

A second useful display is the median shrinkage factor with an interval for each
covariate:

```{r shrinkage-intervals, fig.width=7, fig.height=4, message=FALSE}
library(dplyr)
library(ggplot2)

shrinkage_summary <- posterior::summarise_draws(
    shrinkage_draws,
    median,
    ~ quantile(.x, 0.1),
    ~ quantile(.x, 0.9)
) |>
    dplyr::rename(q10 = `10%`, q90 = `90%`)

ggplot(shrinkage_summary, aes(x = median, y = variable)) +
    geom_errorbar(aes(xmin = q10, xmax = q90), width = 0) +
    geom_point() +
    scale_x_continuous(limits = c(0, 1)) +
    labs(
        x = "Shrinkage factor",
        y = NULL
    )
```

## Interpret the Output

In this simulated example, the design matrix contains `cov_catB`, `cov_catC`,
and `cov_cont`. The data were generated with a small effect for `cov_catB`, a
larger positive effect for `cov_catC`, and a moderate positive effect for
`cov_cont`.

The horseshoe analysis should therefore be interpreted along these lines:

- A shrinkage factor near 1 for `cov_catB`, combined with a coefficient posterior
  close to zero, would indicate that the model sees little evidence for that
  weak covariate.
- Smaller shrinkage factors for `cov_catC` and `cov_cont`, combined with
  coefficient posteriors away from zero, would indicate that these covariates
  escaped the global shrinkage.
- Intermediate shrinkage factors should be treated as uncertainty, not as an
  automatic include/exclude decision.

The practical decision is still scientific. The regularized horseshoe helps rank
and regularize many candidate covariates in a single joint model fit, but final
selection should also consider prior clinical plausibility, multiplicity of
candidate transformations, model diagnostics, and posterior predictive
performance.

## Practical Guidance

The global scale is the main sparsity control. Smaller `scale_global` values
expect fewer relevant covariates; larger values allow more coefficients to
remain away from zero. In applied work, this choice should reflect the number of
candidate covariates and the expected number of non-negligible effects.

The slab scale should be large enough for plausible survival effects but not so
large that implausible log-hazard ratios are effectively unregularized. For
standardized continuous covariates, a `scale_slab` around 2 is often already
wide on the log-hazard scale. For unstandardized covariates, first consider
whether the covariate scale itself should be transformed or standardized.

Finally, the horseshoe prior is a shrinkage prior, not a replacement for model
checking. Always inspect convergence diagnostics, posterior predictive fit, and
sensitivity to reasonable prior choices before using the selected covariates for
scientific conclusions.

## References
