---
title: "Fitting a Custom Longitudinal Model with Covariates"
package: jmpost
format:
  html:
    toc: true
    html-math-method: mathml
link-citations: true
vignette: >
  %\VignetteIndexEntry{Fitting a Custom Longitudinal Model with Covariates}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

## Introduction

This vignette gives a complete implementation of a custom longitudinal model
whose population parameters depend on subject-level covariates. It extends the
Wang, Sung et al. (2009) mixed exponential-decay and linear-growth example from
the [custom longitudinal model vignette](custom-model.html). The general extension
interfaces are described in [Extending jmpost](extending-jmpost.html).

We use the longitudinal model

$$
\begin{align}
y_i(t) &\sim N\left(b_i e^{-s_i t} + g_i t,\ \sigma^2\right), \\
b_i &\sim \operatorname{LogNormal}\left(\eta_{b,i},\ \omega_b\right), \\
s_i &\sim \operatorname{LogNormal}\left(\log(\mu_s),\ \omega_s\right), \\
g_i &\sim \operatorname{LogNormal}\left(\eta_{g,i},\ \omega_g\right), \\
\eta_{b,i} &= \beta_{b,0} + X_{b,i}\beta_b, \\
\eta_{g,i} &= \beta_{g,0} + X_{g,i}\beta_g.
\end{align}
$$

Here, $X_{b,i}$ and $X_{g,i}$ are independently specified design-matrix rows.
Consequently, the baseline and growth distributions can use different
covariates. The intercepts and coefficients are on the log scale, so the
population medians of $b_i$ and $g_i$ are positive.

The survival submodel and derivative link are the same as in the example
without longitudinal covariates:

$$
\log h_i(t) = \log(\lambda_0) +
\alpha\left\{-s_i b_i e^{-s_i t} + g_i\right\}.
$$

The example deliberately implements the covariate handling itself. For a
real application, you would of course first check whether one of `jmpost`'s built-in covariate
models already describes the required trajectory.

```{r}
library(jmpost)
library(ggplot2)
library(dplyr)
```

## Generating Simulated Data

We begin with a simulator so that the fitted arm effects have known target
values. Arm B has a 25% higher median baseline and a 40% higher median growth
rate than Arm A.

```{r}
SimWangCov <- setClass(
    "SimWangCov",
    contains = "SimLongitudinal",
    slots = c(
        times = "numeric",
        mu_baseline = "numeric",
        baseline_arm_ratio = "numeric",
        mu_shrinkage = "numeric",
        mu_growth = "numeric",
        growth_arm_ratio = "numeric",
        omega_baseline = "numeric",
        omega_shrinkage = "numeric",
        omega_growth = "numeric",
        sigma = "numeric",
        link_dsld = "numeric"
    )
)

sampleSubjects.SimWangCov <- function(object, subjects_df) {
    arm_b <- as.numeric(subjects_df$arm == "Arm-B")
    n_subjects <- nrow(subjects_df)

    subjects_df$baseline <- rlnorm(
        n_subjects,
        log(object@mu_baseline) + log(object@baseline_arm_ratio) * arm_b,
        object@omega_baseline
    )
    subjects_df$shrinkage <- rlnorm(
        n_subjects,
        log(object@mu_shrinkage),
        object@omega_shrinkage
    )
    subjects_df$growth <- rlnorm(
        n_subjects,
        log(object@mu_growth) + log(object@growth_arm_ratio) * arm_b,
        object@omega_growth
    )
    subjects_df
}

sampleObservations.SimWangCov <- function(object, times_df) {
    expected <- with(
        times_df,
        baseline * exp(-shrinkage * time) + growth * time
    )
    derivative <- with(
        times_df,
        growth - baseline * shrinkage * exp(-shrinkage * time)
    )

    times_df$sld <- rnorm(nrow(times_df), expected, object@sigma)
    times_df$log_haz_link <- object@link_dsld * derivative
    times_df
}

set.seed(7294)
simulated <- SimJointData(
    design = list(
        SimGroup(25, "Arm-A", "Study-X"),
        SimGroup(25, "Arm-B", "Study-X")
    ),
    survival = SimSurvivalExponential(
        lambda = 365 / 500,
        time_max = 3,
        time_step = 1 / 365,
        lambda_censor = 1 / 30
    ),
    longitudinal = SimWangCov(
        times = c(1, 60, 120, 240, 400, 650, 900) / 365,
        mu_baseline = 60,
        baseline_arm_ratio = 1.25,
        mu_shrinkage = 2,
        mu_growth = 10,
        growth_arm_ratio = 1.4,
        omega_baseline = 0.25,
        omega_shrinkage = 0.25,
        omega_growth = 0.25,
        sigma = 1.5,
        link_dsld = 0.02
    ),
    .silent = TRUE
)

dat_lm <- simulated@longitudinal
dat_os <- simulated@survival

dat_lm |>
    filter(subject %in% head(unique(subject), 8)) |>
    ggplot(aes(time, sld, group = subject, colour = arm)) +
    geom_line(alpha = 0.7) +
    geom_point() +
    labs(x = "Time (years)", y = "Tumour size", colour = "Arm") +
    theme_bw()
```

## Defining the Covariate-Aware Model

A custom covariate model needs to retain its formulas so that the same design
can be reconstructed for fitting and prediction. We therefore add two formula
slots to the `LongitudinalModel` subclass.

```{r}
.WangCovModel <- setClass(
    "WangCovModel",
    contains = "LongitudinalModel",
    slots = c(
        baseline_formula = "formula",
        growth_formula = "formula"
    )
)

WangCovModel <- function(baseline_formula = ~arm,
                         growth_formula = ~arm) {
    stopifnot(
        inherits(baseline_formula, "formula"),
        length(baseline_formula) == 2,
        inherits(growth_formula, "formula"),
        length(growth_formula) == 2
    )

    .WangCovModel(
        LongitudinalModel(
            name = "Wang with Covariates",
            stan = StanModule("custom-cov-model.stan"),
            scaled_variance = FALSE,
            parameters = ParameterList(
                Parameter(
                    name = "log_mu_baseline",
                    prior = prior_normal(log(60), 0.5)
                ),
                Parameter(
                    name = "beta_baseline",
                    prior = prior_normal(0, 0.5),
                    size = "p_baseline_cov"
                ),
                Parameter(
                    name = "mu_shrinkage",
                    prior = set_limits(
                        prior_lognormal(log(2), 0.5),
                        lower = getOption("jmpost.double_eps")
                    )
                ),
                Parameter(
                    name = "log_mu_growth",
                    prior = prior_normal(log(10), 0.5)
                ),
                Parameter(
                    name = "beta_growth",
                    prior = prior_normal(0, 0.5),
                    size = "p_growth_cov"
                ),
                Parameter(
                    name = "sigma_baseline",
                    prior = set_limits(
                        prior_lognormal(log(0.25), 0.5),
                        lower = getOption("jmpost.double_eps")
                    )
                ),
                Parameter(
                    name = "sigma_shrinkage",
                    prior = set_limits(
                        prior_lognormal(log(0.25), 0.5),
                        lower = getOption("jmpost.double_eps")
                    )
                ),
                Parameter(
                    name = "sigma_growth",
                    prior = set_limits(
                        prior_lognormal(log(0.25), 0.5),
                        lower = getOption("jmpost.double_eps")
                    )
                ),
                Parameter(
                    name = "sigma",
                    prior = set_limits(
                        prior_lognormal(log(1.5), 0.5),
                        lower = getOption("jmpost.double_eps")
                    )
                ),
                Parameter(
                    name = "baseline_idv",
                    prior = set_limits(
                        prior_init_only(prior_lognormal(log(60), 0.5)),
                        lower = getOption("jmpost.double_eps")
                    ),
                    size = "n_subjects"
                ),
                Parameter(
                    name = "shrinkage_idv",
                    prior = set_limits(
                        prior_init_only(prior_lognormal(log(2), 0.5)),
                        lower = getOption("jmpost.double_eps")
                    ),
                    size = "n_subjects"
                ),
                Parameter(
                    name = "growth_idv",
                    prior = set_limits(
                        prior_init_only(prior_lognormal(log(10), 0.5)),
                        lower = getOption("jmpost.double_eps")
                    ),
                    size = "n_subjects"
                )
            )
        ),
        baseline_formula = baseline_formula,
        growth_formula = growth_formula
    )
}

longmodel <- WangCovModel(
    baseline_formula = ~arm,
    growth_formula = ~arm
)
```

The corresponding Stan program declares the two model-specific design matrices.
It uses them to form subject-specific log-normal locations and also populates
`long_obvs_log_lik`, which keeps the standard likelihood and `loo` processing
available.

```{r}
#| output: asis
#| echo: false

x <- readLines("./custom-cov-model.stan")
cat(c("\```stan", x, "\```"), sep = "\n")
```

## Supplying Model-Specific Stan Data

`DataLongitudinal()` deliberately continues to use an `outcome ~ time` formula.
Longitudinal covariates are subject-level variables and belong in the data passed
to `DataSubject()`. A `longitudinal_model_stan_data.<class>()` method turns those
variables into the model-specific Stan data.

The helper below removes the usual intercept column because the Stan model has
explicit intercept parameters. Its `reference` argument preserves the fitted
factor levels, contrasts, and column order when the same formula is applied to
new population profiles.

```{r}
covariate_design <- function(formula, data, reference = data) {
    reference_frame <- model.frame(
        formula,
        data = reference,
        na.action = na.fail
    )
    formula_terms <- terms(reference_frame)
    reference_design <- model.matrix(formula_terms, reference_frame)

    data_frame <- model.frame(
        formula_terms,
        data = data,
        xlev = stats::.getXlevels(formula_terms, reference_frame),
        na.action = na.fail
    )
    design <- model.matrix(
        formula_terms,
        data_frame,
        contrasts.arg = attr(reference_design, "contrasts")
    )
    design[, colnames(design) != "(Intercept)", drop = FALSE]
}

longitudinal_model_stan_data.WangCovModel <- function(model, subject) {
    subject_data <- as.data.frame(subject)
    baseline_design <- covariate_design(
        model@baseline_formula,
        subject_data
    )
    growth_design <- covariate_design(
        model@growth_formula,
        subject_data
    )

    list(
        p_baseline_cov = ncol(baseline_design),
        baseline_cov_design = baseline_design,
        p_growth_cov = ncol(growth_design),
        growth_cov_design = growth_design
    )
}
```

The method name must end in the exact S4 class name. `jmpost` calls it while
converting `DataJoint` to Stan data, so users do not need to construct or append
the matrices themselves.

## Defining the Link Function

The link setup is unchanged by the population covariates: downstream link
functions consume the realized subject parameters.

```{r}
enableLink.WangCovModel <- function(object, ...) {
    object@stan <- merge(
        object@stan,
        StanModule("custom-cov-model-enable-link.stan")
    )
    object
}

link <- LinkComponent(
    stan = StanModule("custom-cov-model-dsld.stan"),
    prior = prior_normal(0, 0.5),
    key = "link_dsld"
)
```

```{r}
#| output: asis
#| echo: false

x <- readLines("./custom-cov-model-enable-link.stan")
cat(c("\```stan", x, "\```"), sep = "\n")
```

```{r}
#| output: asis
#| echo: false

x <- readLines("./custom-cov-model-dsld.stan")
cat(c("\```stan", x, "\```"), sep = "\n")
```

## Sampling the Joint Model

The subject data contains `arm`, which is used by both longitudinal formulas.
The survival formula has no additional covariates in this simple example.

```{r}
#| output: false

model <- JointModel(
    longitudinal = longmodel,
    survival = SurvivalExponential(lambda = prior_gamma(1, 1)),
    link = link
)

joint_data <- DataJoint(
    subject = DataSubject(
        data = dat_os,
        subject = "subject",
        arm = "arm",
        study = "study"
    ),
    survival = DataSurvival(
        data = dat_os,
        formula = Surv(time, event) ~ 1
    ),
    longitudinal = DataLongitudinal(
        data = dat_lm,
        formula = sld ~ time
    )
)

model_samples <- sampleStanModel(
    model,
    data = joint_data,
    iter_warmup = 400,
    iter_sampling = 400,
    chains = 1,
    parallel_chains = 1,
    refresh = 0,
    adapt_delta = 0.9
)
```

```{r}
fit <- cmdstanr::as.CmdStanMCMC(model_samples)
fit$summary(c(
    "log_mu_baseline",
    "beta_baseline",
    "log_mu_growth",
    "beta_growth",
    "mu_shrinkage",
    "sigma",
    "link_dsld"
))
```

The true arm coefficients are `log(1.25)` for baseline and `log(1.4)` for
growth. The short, single-chain run keeps vignette execution manageable; use
multiple chains and sampling iterations, and check all diagnostics, for a real application.

## Enabling Downstream Processing

Individual predictions only need each subject's fitted parameters (implemented with `QuantityGeneratorSubject`). Population
predictions additionally need design matrices evaluated at the requested
covariate profiles (implemented with `QuantityGeneratorPopulation`). We keep these cases separate so that Stan code never refers
to population-only data while generating subject-level quantities.

```{r}
enableGQ.WangCovModel <- function(object,
                                  generator = NULL,
                                  type = NULL,
                                  ...) {
    if (!identical(type, "longitudinal")) {
        return(StanModule())
    }

    result <- StanModule("custom-cov-model-gq.stan")
    if (is(generator, "QuantityGeneratorSubject")) {
        result <- merge(
            result,
            StanModule("custom-cov-model-gq-individual.stan")
        )
    } else if (is(generator, "QuantityGeneratorPopulation")) {
        result <- merge(
            result,
            StanModule("custom-cov-model-gq-population.stan")
        )
    }
    result
}

required_longitudinal_covs.WangCovModel <- function(object, ...) {
    unique(c(
        all.vars(object@baseline_formula),
        all.vars(object@growth_formula)
    ))
}

gq_population_stan_data.WangCovModel <- function(
  object,
  model,
  data = NULL,
  ...
) {
    result <- list(
        declarations = paste(
            "matrix[gq_n_quant, p_baseline_cov] gq_baseline_cov_design;",
            "matrix[gq_n_quant, p_growth_cov] gq_growth_cov_design;",
            sep = "\n"
        ),
        data = list()
    )
    if (is.null(data)) {
        return(result)
    }

    stopifnot(!is.null(object@newdata))
    subject_data <- as.data.frame(data@subject)
    result$data <- list(
        gq_baseline_cov_design = covariate_design(
            model@baseline_formula,
            object@newdata,
            reference = subject_data
        ),
        gq_growth_cov_design = covariate_design(
            model@growth_formula,
            object@newdata,
            reference = subject_data
        )
    )
    result
}

getRandomEffectsNames.WangCovModel <- function(object, ...) {
    c(
        baseline = "baseline_idv",
        shrinkage = "shrinkage_idv",
        growth = "growth_idv"
    )
}
```

The shared prediction function is:

```{r}
#| output: asis
#| echo: false

x <- readLines("./custom-cov-model-gq.stan")
cat(c("\```stan", x, "\```"), sep = "\n")
```

The subject-level and covariate-aware population parameter blocks are:

```{r}
#| output: asis
#| echo: false

x <- c(
    readLines("./custom-cov-model-gq-individual.stan"),
    "",
    readLines("./custom-cov-model-gq-population.stan")
)
cat(c("\```stan", x, "\```"), sep = "\n")
```

We can now run the downstream methods. First, extract and summarize the three
subject-level random effects:

```{r}
random_effects <- LongitudinalRandomEffects(model_samples)
head(summary(random_effects))
```

Next, generate fitted trajectories for selected observed subjects:

```{r}
selected_subjects <- head(dat_os$subject, 4)
long_quantities_idv <- LongitudinalQuantities(
    model_samples,
    grid = GridFixed(subjects = selected_subjects)
)
autoplot(long_quantities_idv)
```

Finally, compare population trajectories for the two arm profiles. Each row of
`newdata` is a complete subject profile. It must include `study` and `arm` for
the standard grouping indexes as well as every variable returned by
`required_longitudinal_covs()`.

```{r}
population_profiles <- data.frame(
    study = "Study-X",
    arm = c("Arm-A", "Arm-B")
)

long_quantities_pop <- LongitudinalQuantities(
    model_samples,
    grid = GridPopulation(
        times = seq(0, 2.5, length.out = 51),
        newdata = population_profiles
    )
)

head(summary(long_quantities_pop))
autoplot(long_quantities_pop)
```

This last call exercises the complete covariate-aware path: `GridPopulation()`
retains the profiles, `required_longitudinal_covs()` validates them,
`gq_population_stan_data()` mirrors the fitted design matrices, and the custom
generated-quantities Stan block evaluates the covariate-dependent population
parameters.
