---
title: "Model Fitting"
package: jmpost
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Fitting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
library(jmpost)
```

Here we describe the basic steps of joint model fitting with `{jmpost}`.

# Model Specification

Each `JointModel` needs to be specified with three parts:

1. `longitudinal`: The model for the longitudinal outcomes.
1. `survival`: The model for the survival outcomes.
1. `link`: The link that specifies how the `longitudinal` model parameters enter the
  `survival` model.

## Default Options

Let's first specify a very simple joint model with:

1. A random slope model for the longitudinal outcome.
1. A Weibull proportional hazards model for the survival outcome.
1. The link between the two models is that the random slope from the longitudinal model
   enters as a product with a link coefficient into the linear predictor of the 
   survival model.
   
```{r simple_model}
simple_model <- JointModel(
    longitudinal = LongitudinalRandomSlope(),
    survival = SurvivalWeibullPH(),
    link = link_dsld()
)
```

Note that here we use all the default options for the two models and the link,
in particular the prior distributions and the initial values in the MCMC chain
for the parameters are automatically chosen. We can see this 
from the arguments of the constructors (or from the help page):

```{r constructor_args}
args(LongitudinalRandomSlope)
```

So here we see that the Longitudinal Random Slope model has 4 parameters that we can
define a prior for.

## Specifying Priors

We can alternatively also specify the prior distributions
for the parameters manually. This is important in practice to obtain a meaningful
model specification and hence converging MCMC chains that allow to estimate the 
posterior distributions.

For the random slope model for the longitudinal outcome, we can e.g. say:

```{r random_slope_prior_inits}
random_slope_model <- LongitudinalRandomSlope(
    intercept = prior_normal(40, 5),
    slope_mu = prior_normal(10, 2)
)
```

This sets the prior for the `intercept` to be a $N(40, 5)$ distribution and
the prior for the `slope_mu` parameter to be a $N(10, 2)$ distribution.

## Separate Models

It is also possible to not link the longitudinal and the survival models, by using
the special `link_none()` link specification. For example, 

```{r simple_model_no_link}
simple_model_no_link <- JointModel(
    longitudinal = LongitudinalRandomSlope(),
    survival = SurvivalWeibullPH(),
    link = link_none()
)
```

would allow to fit the two models separately, but in the same MCMC chain.

## Single Models

It is also possible to specify only the longitudinal or the survival model.
Then these can be estimated on their own with separate MCMC chains.

```{r single_models}
single_longitudinal <- JointModel(
    longitudinal = LongitudinalRandomSlope()
)
single_survival <- JointModel(
    survival = SurvivalWeibullPH()
)
```

# Data Preparation

Before we can fit the models, we need to prepare the data in the right format.

## Simulating Data

Here we start from a simulated data set. 

- We assign 50 patients each to the two treatment arms.
- We use a time grid from 1 to 2000, e.g. specifying the days after randomization.
- We use an exponentially distributed censoring time with mean of 9000 days.
- We use a categorical covariate with three levels A, B and C in the overall survival model, drawn uniformly from the three levels. (Note that this is hardcoded at the moment, so the levels need to be A, B, C.)
- We use another continuous covariate in the overall survival model generated from a standard normal distribution, with coefficient 0.3.
- For the longitudinal outcome, we draw the values from a random slope model with the given parameters.
- For the survival outcome, we draw the true value from a Weibull model. Note that it is fairly easy to put here another choice, you just need to specify a function of `time` returning the log baseline hazard under the given survival model.

So let's run the code for that:

```{r simulate_data}
set.seed(129)
sim_data <- SimJointData(
    design = list(
        SimGroup(50, "Arm-A", "Study-X"),
        SimGroup(50, "Arm-B", "Study-X")
    ),
    longitudinal = SimLongitudinalRandomSlope(
        times = c(1, 50, 100, 150, 200, 250, 300),
        intercept = 30,
        slope_mu = c(1, 2),
        slope_sigma = 0.2,
        sigma = 20,
        link_dsld = 0.1
    ),
    survival = SimSurvivalWeibullPH(
        lambda = 1 / 300,
        gamma = 0.97,
        time_max = 2000,
        time_step = 1,
        lambda_cen = 1 / 9000,
        beta_cat = c(
            "A" = 0,
            "B" = -0.1,
            "C" = 0.5
        ),
        beta_cont = 0.3
    )
)
```

We might get a message here that a few patients did not die before the day 2000, but 
this is not of concern. Basically it just gives us a feeling of how many survival
events are included in the data set.

## Formatting Data

Next we bring the data into the right format.

We start with extracting data into individual data sets, and then reducing the
longitudinal data to specific time points.

```{r individual_data}
os_data <- sim_data@survival
long_data <- sim_data@longitudinal
```

Let's have a quick look at the format:

The survival data has:

- patient ID
- time point
- continuous covariate value
- categorical covariate level
- event indicator (1 for observed, 0 for censored)
- study ID
- treatment arm

```{r os_data_structure}
head(os_data)
```

The longitudinal data has:

- patient ID
- time point
- sum of longest diameters (SLD)
- study ID
- treatment arm
- observation flag

```{r lm_data_structure}
head(long_data)
```

Finally, we wrap these in the data formatting functions. Here the mapping of the
column names to the required variables happens. This means that in our applications
we don't have to use the same variable names as seen above, but we can use custom 
names and then apply the mapping here.

```{r data_joint}
joint_data <- DataJoint(
    subject = DataSubject(
        data = os_data,
        subject = "pt",
        arm = "arm",
        study = "study"
    ),
    survival = DataSurvival(
        data = os_data,
        formula = Surv(time, event) ~ cov_cat + cov_cont
    ),
    longitudinal = DataLongitudinal(
        data = long_data,
        formula = sld ~ time,
        threshold = 5
    )
)
```

# Model Fitting

Now let's have a look how we can fit the (joint) models.

## Debugging Stan Code

It is always possible to read out the Stan code that is contained in the
`JointModel` object, using `write_stan()`:

```{r debug_stan}
tmp <- tempfile()
write_stan(simple_model, file_path = tmp)
first_part <- head(readLines(tmp), 10)
cat(paste(first_part, collapse = "\n"))
```

## Sampling Parameters

Finally, `sampleStanModel()` is kicking off the MCMC sampler via `cmdstanr` in
the backend. Note that in practice you would increase the number of 
warm-up and sampling iterations.

```{r sampling}
mcmc_results <- sampleStanModel(
    simple_model,
    data = joint_data,
    iter_sampling = 500,
    iter_warmup = 500,
    chains = 1,
    parallel_chains = 1
)
```



## Convergence checks

After the sampling finishes, we can inspect the parameter distributions.
This is using the `cmdstanr` functions, because the `results` element is of class
`CmdStanMCMC`.

```{r inspect_distributions}
vars <- c(
    "lm_rs_intercept",
    "lm_rs_slope_mu",
    "lm_rs_slope_sigma",
    "lm_rs_sigma",
    "link_dsld",
    "sm_weibull_ph_lambda",
    "sm_weibull_ph_gamma",
    "beta_os_cov"
)
mcmc_results@results$summary(vars)
```

We can already see here from the `rhat` statistic whether the MCMC sampler
converged - values close to 1 indicate convergence, while values larger than
1 indicate divergence.

In general, convergence is sensitive to the choice of:

- Priors
- Initial values
- Sufficient warm-up iterations

If you don't achieve convergence, then play around with different choices
of the above.

# Plotting

We can now proceed towards investigating the results of the MCMC chain with 
plots.

## Trace plots

Using the `draws()` method on the `results` element, we can extract the samples
in a format that is understood by `bayesplot`. We can e.g. look 
at some simple trace plots:

```{r fig.width=6}
vars_draws <- mcmc_results@results$draws(vars)

library(bayesplot)
mcmc_trace(vars_draws)
```

## Longitudinal fit plots

Using the `longitudinal()` method we can extract the longitudinal fit
samples from the result, and then plot them for all patients or those
that we are interested in. For illustration, we will plot here the first
10 patients:

```{r fig.width=6}
selected_patients <- head(os_data$pt, 10)
long_quantities <- LongitudinalQuantities(
    mcmc_results,
    groups = selected_patients
)
as.data.frame(long_quantities) |> head()
summary(long_quantities) |> head()
autoplot(long_quantities)
```

## Survival fit plots

And using the `SurvivalQuantities()` method we can do the same for the estimated survival
functions.

```{r fig.width=6}
surv_quantities <- SurvivalQuantities(
    mcmc_results,
    groups = selected_patients
)
as.data.frame(surv_quantities) |> head()
summary(surv_quantities) |> head()
autoplot(surv_quantities)
```

We can also aggregate the estimated survival curves from groups of patients, 
using the corresponding method.

```{r, warning=FALSE, fig.width=6}
groups <- split(os_data$pt, os_data$arm)
surv_quantities <- SurvivalQuantities(
    mcmc_results,
    groups = groups
)
autoplot(surv_quantities, add_km = TRUE)
```



## Brier Score

The `brierScore()` method can be used to extract the Brier Scores (predictive performance measure)
from our `SurvivalQuantities` object.

```{R}
prediction_times <- c(1, 50, 100, 400, 800)
sq <- SurvivalQuantities(
    mcmc_results,
    time_grid = prediction_times,
    type = "surv"
)
brierScore(sq)
```


## Initial Values

By default `jmpost` will set the initial values for all parameters to be a random value
drawn from the prior distribution that has been shrunk towards the
mean of said distribution e.g. for a `prior_normal(4, 2)`
the initial value for each chain will be: 
```
4 * shrinkage_factor + rnorm(1, 4, 2) * (1 - shrinkage_factor)
```

Note that the shrinkage factor is set to 0.5 by default and can be changed via the
`jmpost.prior_shrinkage` option e.g.

```
options("jmpost.prior_shrinkage" = 0.7)
```

If you wish to manually specify the initial values you can do so via the `initialValues()` function.
For example:

```r
joint_model <- JointModel(
    longitudinal = LongitudinalRandomSlope(),
    survival = SurvivalExponential(),
    link = link_none()
)
initial_values <- initialValues(joint_model, n_chains = 2)

initial_values[[1]]$lm_rs_intercept <- 0.2
initial_values[[2]]$lm_rs_intercept <- 0.3

mcmc_results <- sampleStanModel(
    joint_model,
    data = DataJoint(...),
    init = initial_values,
    chains = 2
)
```

Note the following:
-  `initialValues()` will return a list of lists where each sublist contains the initial
values for the corresponding chain index.
- `initialValues()` by default just returns 1 initial value for each parameter; the `sampleStanModel()`
function will then broadcast this number as many times as required e.g. if you have 3 covariates
in your survival model then the initial value for the Beta coeficient will be repeated 3 times. If
however you want to specify individual initial values for each covariate you can do so by passing
in a vector of the same length as the number of covariates.



