---
title: "Simulating Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(jmpost)
```

The `jmpost` package includes data simulation functionality for the included joint models.
The data simulation is based on specifying the longitudinal model and the survival model including link parameters.


## Example


```{r simulate_data}
set.seed(129)
sim_data <- SimJointData(
    design = list(
        SimGroup(50, "Arm-A", "Study-X"),
        SimGroup(50, "Arm-B", "Study-X")
    ),
    longitudinal = SimLongitudinalSteinFojo(
        times = c(-2, 1,30, 90, 150, 210),
        mu_g = log(c(0.005, 0.005)),
        mu_s = log(c(0.06, 0.007)),
        omega_g = 0.3,
        link_dsld = 0.1,
    ),
    survival = SimSurvivalWeibullPH(
        lambda = 1 / 300,
        gamma = 0.97,
        time_max = 2000,
        time_step = 1,
        lambda_cen = 1 / 9000,
        beta_cat = c(
            "A" = 0,
            "B" = -0.1,
            "C" = 0.5
        ),
        beta_cont = 0.3
    )
)
```

This object has `survival` and `longitudinal` components.
```{r}
head(sim_data@survival)
head(sim_data@longitudinal)
```

We can see the trajectory of the tumour size.

```{r}
library(ggplot2)
ggplot(
    sim_data@longitudinal,
    aes(x = time, y = sld, group = subject)) +
    geom_line(alpha = 0.1) +
    geom_point() +
    facet_wrap(~arm)

```

We can also visualise the Kaplan-Meier survival curves.

```{r}
library(survival)
plot(
    survfit(Surv(time, event) ~ arm, data = sim_data@survival),
    col = 1:2,
    lwd = 2,
    main = "Overall Survival"
)
legend("topright", col = 1:2, lwd = 2, legend = c("Arm-A", "Arm-B"))
```

## Administrative Cut Off

We may wish to have a fixed maximum study duration and remove any simulated values after that time.
This can be specified as a single fixed value or as a vector if we additionally wish to consider
enrollment time.

```{r}
sim_data_end <- cut_data(sim_data, 700)
plot(
    survfit(Surv(time, event) ~ arm, data = sim_data_end@survival),
    col = 1:2,
    lwd = 2,
    main = "Overall Survival (truncated at time=700)"
)
```


## Progression

We might also be interested in time to progression defined as an increase in tumour size over a threshold.
The threshold is based on the minimum observed SLD up to the given time and the relative and absolute growth.
For example we might require the SLD to increase at least 20% and at least 5mm.
We don't include any observations before time 0 in the calculation of the minimum and we don't observe any SLD values 
after progression.
```{r}
sim_data_pd <- add_pfs(
    sim_data_end,
    relative_threshold = 1.2,
    absolute_threshold = 5,
    from_time = 0,
    observed_after = FALSE
)
```

Let's look at the plots again up to progression.

```{r}
ggplot(
    sim_data_pd@longitudinal |> dplyr::filter(observed), # here we only include observed values
    aes(x = time, y = sld, group = subject)) +
    geom_line(alpha = 0.1) +
    geom_point() +
    facet_wrap(~arm)

```

```{r}
plot(
    survfit(Surv(pfs_time, pfs_event) ~ arm, data = sim_data_pd@survival),
    col = 1:2,
    lwd = 2,
    main = "Progression Free Survival"
)
legend("topright", col = 1:2, lwd = 2, legend = c("Arm-A", "Arm-B"))
```

## Model Fitting from Simulated Data

As already descried in the quickstart vignette, we can use this data to fit a joint model.

```{r}
os_data <- sim_data_end@survival
long_data <- sim_data_end@longitudinal
joint_data <- DataJoint(
    subject = DataSubject(
        data = os_data,
        subject = "subject",
        arm = "arm",
        study = "study"
    ),
    survival = DataSurvival(
        data = os_data,
        formula = Surv(time, event) ~ cov_cat + cov_cont
    ),
    longitudinal = DataLongitudinal(
        data = long_data,
        formula = sld ~ time,
        threshold = 5
    )
)



sf_model <- JointModel(
    longitudinal = LongitudinalSteinFojo(
        mu_bsld = prior_normal(log(60), 0.2),
        mu_ks = prior_normal(-3, 0.4),
        mu_kg = prior_normal(log(0.005), 0.3),
        centred = TRUE
    ),
    survival = SurvivalWeibullPH(),
    link = linkDSLD()
)

set.seed(2026)
mcmc_results <- sampleStanModel(
    sf_model,
    data = joint_data,
    iter_sampling = 500,
    iter_warmup = 500,
    chains = 4,
    parallel_chains = 4,
    step_size = 0.05
)
mcmc_results@results
```

## Simulating from a Fitted Model

The package also have the functionality to generate new data based on the fitted model.
We use the [simulate.JointModelSamples] function.

```{r}
?simulate.JointModelSamples
new_model_data <- simulate(
    mcmc_results,
    times = c(-2, 1, 30, 60, 90, 120, 150, 180),
    jitter_var = c(0, 2)
)
```

For this simulation we set 

- the times we wish to have longitudinal observations
- variance for jitter around those observation times
- time_max and time_step for numerical integration
- lambda parameter for exponential censoring times
- scaled_variance which should be set to the same as the fitted model. This is described more in 
  the statistical specifications vignette.

Now we can inspect the new simulated data.

```{r}
head(new_model_data@longitudinal)
head(new_model_data@survival)
```


```{r}
ggplot(
    dplyr::bind_rows(
        "model sim" = new_model_data@longitudinal,
        "original data" = sim_data@longitudinal,
        .id = "sim"
    ),
    aes(x = time, y = sld, group = subject)) +
    geom_line(alpha = 0.1) +
    geom_point() +
    facet_wrap(~arm + sim) +
    ggtitle("Original Data versus Model Simulated Data")

```

```{r}
plot(
    survfit(Surv(time, event) ~ arm, data = new_model_data@survival),
    col = 1:2,
    lwd = 2,
    main = "Overall Survival (simulated from model)"
)
lines(
    survfit(Surv(time, event) ~ arm, data = sim_data@survival),
    col = 1:2,
    lwd = 2,
    lty = 2
    )
legend("topright", col = c(1, 2, 1, 2), lwd = 2, lty = c(1, 1, 2, 2),
       legend = c("Arm-A from model", "Arm-B from model", "Arm-A original", "Arm-B original"))
```

