---
title: "{jmpost} documentation"
subtitle: "01. Survival models"
package: jmpost
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Fitting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

## Scope of this document

This document describes how to analyze survival data using parametric
survival models implemented in the **jmpost** R package. It is organized
in 4 sections:

1.  Introduction

2.  Survival model framework

3.  Fitting a simple survival model in **jmpost**

4.  Comparing with **flexsurv** and **survstan** outputs

## 1. Introduction

Survival (or time-to-event, or failure time) analysis is concerned with
the study of survival times and factors influencing them.

There are 3 common approaches to modelling survival data:

1.  Models of the instantaneous rate of the event (known as the hazard)
    as a function of time taken as a continuous variable. This includes
    the well-known class of proportional hazards (PH) regression models.

2.  Models the event time itself, again considering time as a continuous
    variable, including the accelerated failure time (AFT) models.

3.  Models of the occurrence of event when time is taken as a discrete
    variable, specifically the additive hazard regression (AHR) model.

In the current version of **jmpost**, we focus on approach 1., and
specifically on PH models, implemented in a Bayesian framework.

Proportional hazard (PH) models play an important role in analyzing
survival and rely on a key assumption: the hazard ratio (HR) comparing
any two levels of a covariate is constant over time. The Cox PH model is
most popular for two reasons: (a) no assumption is required about the
probability distribution of survival times, and (b) it fits the data
reasonably well no matter which parametric model is appropriate.

In contrast, distributional assumption is required for a fully
parametric PH model (Kalbfleisch and Prentice 2002 :: The Statistical
Analysis of Failure Time Data. John Wiley & Sons, New Jersey; Lawless
2002 :: Statistical Models and Methods for Lifetime Data. John Wiley &
Sons, New Jersey (2002)). When the distribution is selected wisely,
parametric models lead to more efficient (i.e. smaller standard errors)
estimates than Cox's model (Collett 2003 :: Modelling Survival Data in
Medical Research. Chapman and Hall/CRC, Florida).

Moreover, the use of Cox PH in joint modeling of time-to-event and
longitudinal data (Wulfsohn and Tsiatis 1997 :: A joint model for
survival and longitudinal data measured with errror. Biometrics. 53,
330--339) usually leads to an underestimation of the parameter estimates
standard errors (Hsieh et al. 2006 :: Joint modeling of survival and
longitudinal data: Likelihood approach revisited. Biometrics. 62,
1037--1043; Rizopoulos 2012 :: Joint Models for Longitudinal and
Time-to-Event Data With Applications in R. Chapman and Hall/CRC,
Florida). Therefore, most methods for joint modeling are based on
parametric response distributions (Hwang and Pennell 2014 ::
Semiparametric bayesian joint modeling of a binary and continuous
outcome with applications in toxicological risk assessment. Stat. Med.
33, 1162--1175).

Bayesian inference offers the opportunity 

- to make probability statements about parameters, 

- to handle hierarchical structures naturally, 

- to have good statistical properties when dealing with small samples, and 

- to quantify uncertainty in predicted quantities easily. 

All these properties are convenient for joint models, but also
for survival models taken separately.

**jmpost** is written in Stan, a high-level language (itself written in
a C++) for Bayesian modeling and inference that primarily uses a version
of the No-U-Turn sampler (NUTS) (Hoffman & Gelman, 2014 :: Hoffman, M.
D., & Gelman, A. (2014). The No-U-turn sampler: Adaptively setting path
lengths in Hamiltonian Monte Carlo. Journal of Machine Learning
Research, 15(47), 1593-1623) to obtain posterior samples given a
user-specified model and data.

Although Stan is an extremely flexible and powerful software for
statistical modelling, it is neither user-friendly nor easy to learn. To
remove this hurdle, we propose **jmpost** as a layer on top of Stan to
facilitate the encoding of PH models, either stand-alone, or integrated
in joint models.

## 2. Survival model framework

Let $T^*_i$ denote a continuous non-negative random variable
representing survival time for individual $i$ (with $i=1, ..., n$).

In reality, the event time $T^*_i$ may not be observed due to
right-censoring; for instance, the event may occur after the end of the
clinical study, denoted $\tau_f$. We define the outcome as the pair:
$T_i$ being the observed event or censoring time (with $T_i>0$) and
$d_i \in \{0,1\}$, an indicator taking value 0 in case of right
censoring (i.e. $T^*_i>\tau_f$), or value 1 if the event is observed
(i.e. $T^*_i=T_i$).

Three key quantities form the likelihood function for survival models:
the hazard rate, the cumulative hazard, and the survival probability.

The survival function $S(t, \psi_i)$ gives the probability that the
event happens to individual $i$ after time t, where $\psi_i$ represent
the individual parameters. 
$$
S(t, \psi_i) = P(T_i>t; \psi_i)
$$

The hazard function $h(t, \psi_i)$ is the instantaneous rate of occurrence of the event for individual $i$ at time $t$, given that the event has not already occurred. Mathematically, it is defined as: 
$$
h(t, \psi_i) = \displaystyle \lim_{dt \to 0} \displaystyle \frac{P(t \le T_i < t+dt \mid T_i > t)}{dt}
$$ 
where $dt$ is a small time interval.

This is equivalent to: 
$$
h(t, \psi_i)= - \frac{d}{dt} \log S(t, \psi_i)
$$ 
Knowing the hazard function gives access to the survival function.

The cumulative hazard is defined as: 
$$
H(a, b, \psi_i)=\int_{a}^{b} h(t, \psi_i) \; dt
$$ 

which implies that: 
$$
S(t, \psi_i) = \exp(-H(0, t, \psi_i))
$$

The log-likelihood function in presence of right-censoring measures the risk of experiencing the event at any time point $t$, conditionally on that individual $i$ has in fact survival until that time point; if s/he has not survived, then the probability of experiencing the event again is 0: 
$$
log \mathcal{L}=\sum_{i = 1}^{n} \left[ d_i \log h(t, \psi_i) + \log S(t, \psi_i) \right]
$$

When formulating a parametric survival model, we need to specify the form of the probability distribution assumed to describe the underlying uncertainty in the observed times.

In general terms, we can specify the vector of relevant parameters as $\theta = (μ(x), \alpha(x))$. In this notation, consistent with **flexsurv**, we consider: 

- a vector of potential covariates $x$ (e.g., treatment group, age, or sex), 

- a *location* parameter $\mu(x)$, which indicates the *mean* or the *scale* of the probability distribution, 

- and a (set of) ancillary parameters $\alpha(x)$, which describe the *shape* or *variance* of the distribution. 

While it is possible for both $\mu$ and $\alpha$ to explicitly depend on the covariates $x$, usually the formulation is simplified to assume that these only affect directly the location parameter.

To parameterize the effect of covariates on the location parameter, a generalized linear formulation is employed:
$$
g(\mu_i)=\beta_0+\sum_{k = 1}^{K} \beta_k x_ik
$$
The function $g(.)$ is typically the logarithm, but alternative transformations such as square-root or Box-Cox could also be considered.

We model the hazard of event for individual $i$ at time $t$ using the regression model: 
$$
h(t, \psi_i)=h_0(t) \exp(\beta_0+\sum_{k = 1}^{K} \beta_k x_{ik})
$$
where $h_0(t)$ is the baseline hazard (i.e. the hazard for an individual with the covariates set equals to zero) at time $t$, and the terms in $exp(.)$ denotes the linear predictors evaluated for individual $i$ at time $t$.

The following distributional assumptions for the baseline hazard are currently implemented in the **jmpost** package:

1- Exponential

2- Weibull

3- Log-logistic

The following table shows the location and ancillary parameter with their default priors.

```{r setup, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
library(flexsurv)
```

```{r distributiontable, echo=FALSE}
ztab <- data.frame(
    Dist = c("\\( \\ T_i \\sim \\exp(\\mu_i) \\)", 
             "\\( \\ T_i \\sim Weibull(\\mu_i, \\alpha) \\)",
             "\\( \\ T_i \\sim logLogistic(\\mu_i, \\alpha) \\)"),
    Loca = c("\\( \\ Rate: \\mu_i=\\exp(\\beta_0+\\sum_{k = 1}^{K} \\beta_k x_{ik}) \\)", 
             "\\( \\ Scale: \\mu_i= (idem) \\)",
             "\\( \\ Rate: \\mu_i= (idem) \\)"),
    Shap = c("\\( NA \\)",
             "\\( \\ Shape: \\alpha \\sim \\gamma(0.1, 0.1) \\)", 
             "\\( \\ Shape: \\alpha \\sim \\gamma(0.1, 0.1) \\)")
)

# Create table
knitr::kable(ztab, 
    escape = F, 
    caption = 'Table example', 
    col.names = c("Distribution", "Location (or Scale)", "Shape"))
```
 
The models implemented in **jmpost** all have a PH parameterization.


## 3. Fitting a simple survival model in jmpost

### 3.1. Call

In the following, we run a simple example to present the features of **jmpost**.

The 'bc' dataset from [**flexsurv**](https://chjackson.github.io/flexsurv/) is used. It contains data for 686 patients with primary node positive breast cancer. The right-censored time-to-event variable `recyrs` represents the time (in years) of death or cancer recurrence when `censrec` is 1, or right censoring when `censrec` is 0. The covariate `group` is a factor representing a prognostic score, with 3 levels: "Good" (the reference), "Medium" and "Poor".

Showing the first 6 and the last 6 rows, the dataset looks like this:
```{r, eval=TRUE, echo=TRUE}
rbind(head(bc), tail(bc))
```

The call to fit an exponential PH model in **jmpost** would be as follow:

```{r, eval=FALSE, echo=TRUE}
jm<-JointModel(survival=
        SurvivalExponential(lambda=prior_lognormal(log(0.06), 1)))

bc1<-bc %>%
  mutate(ID=as.character(1:n()), study=1)
jdat<-DataJoint(
  subject = DataSubject(data=bc1, subject="ID", arm="group", study="study"),
  survival = DataSurvival(data=bc1, formula=Surv(recyrs, censrec)~group)
)

mp<-sampleStanModel(jm, data=jdat, iter_warmup=4000, iter_sampling=1000, chains=4)
vars<-c("sm_exp_lambda", "beta_os_cov")
mp@results$summary(vars)

```

Details about the parameters and functions are available in the [Model Fitting](https://genentech.github.io/jmpost/main/articles/model_fitting.html).

### 3.2 Default outputs

The default output reports statistics for one or more sampler chains over all model parameters and quantities of interest. 
The statistics reported include both summary statistics of the estimates and diagnostic statistics on the sampler chains, reported in the following order: 
- mean: sample mean, 
- median: sample median,
- sd: sample standard deviation,
- q5, q95: 5th and 95th percentiles of the posterior distribution,
- rhat: 


```{r}

#' Convert the samples in a df
#' ------------------------------------------------------------\
mp@results$summary(vars)

my_fit_df <- posterior::as_draws_df(mp@results)
my_pars<-c("sm_exp_lambda", "beta_os_cov[1]", "beta_os_cov[2]")


#' Summarize the posterior samples
#' ------------------------------------------------------------\
#' Using {posterior}
posterior::summarise_draws(my_fit_df)

```

 
### 3.3 Goodness-of-fit evaluation


### 3.4 Observed vs. fitted survival



<!--

## Comparing {flexsurv}, {survstan} and {jmpost}

### Data preparation

### Flexsurv implementation

The details of the exponential model implementation in {flexsurv} are
provided in the reference article
([DOI](https://www.jstatsoft.org/article/view/v070i08)). The exponential
distribution has a single parameter and only supports a hazard that is
constant over time. The hazard is equal to the rate parameter.

```{r, eval=FALSE, echo=TRUE}
flexsurv.exp<-flexsurv::flexsurvreg(Surv(recyrs, censrec)~group, data=bc, dist="exp")
flexsurv.exp
```

The same model in {survstan} would be implemented as follow:

```{r, eval=FALSE, echo=TRUE}
survstan.exp<-survstan::phreg(Surv(recyrs, censrec)~group, data=bc, dist="exponential")
summary(survstan.exp)

```


As shown in the below table, these various model calls provide
consistent outputs. More details on the {jmpost} outputs are provided in
the separate vignette ABC123.

```{r, file="expon-ph-benchmark.R", echo=FALSE}
library(knitr)
library(kableExtra)

# Create table
altogether %>%
  kable(escape = F) %>%
  kable_styling()
```

Both `jmpost` and `survstan` implement these analysis in Stan, a
probabilistic programming language supporting Bayesian approach.

--> 
