---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Design for the survival fit plots

Now let's try to wrap our head around how we can produce nice survival
fit plots from our joint or single survival model fit.

## Example

Let's start with a simple example - the simplest is when we just have the
survival model:

```{r}
single_survival <- JointModel(
    survival = SurvivalLogLogistic()
)
set.seed(321)
sim_data <- simulate_joint_data(
    lm_fun = sim_lm_random_slope(),
    os_fun = sim_os_exponential(1 / 10),
    lambda_cen = 1 / 300
)
os_data <- sim_data$os
long_data <- sim_data$lm |>
    dplyr::filter(time %in% c(1, 50, 100, 150, 200, 250, 300)) |>
    dplyr::arrange(time, pt)
joint_data <- DataJoint(
    survival = DataSurvival(
        data = os_data,
        formula = Surv(time, event) ~ cov_cat + cov_cont,
        subject = "pt",
        arm = "arm",
        study = "study"
    ),
    longitudinal = DataLongitudinal(
        data = long_data,
        formula = sld ~ time,
        subject = "pt",
        threshold = 5
    )
)
mcmc_results <- sampleStanModel(
    single_survival,
    data = joint_data,
    iter_sampling = 500,
    iter_warmup = 500,
    chains = 1,
    parallel_chains = 1,
    exe_file = file.path("local", "single_survival")
)
mcmc_results
```

It can be instructive to look at the Stan code:

```{r, eval = FALSE}
tmp <- tempfile()
write_stan(single_survival, file_path = tmp)
file.edit(tmp)
```

## Obtaining survival fit samples for single patients

So how do we now obtain patient specific samples of the fitted survival model?

### Generate samples in Stan

We want to generate the samples in Stan already as part of the generated quantities
block. Note that it is not sufficient to just have them at the observed survival time
points, because then we cannot plot nice functional fits afterwards. So we need this at a 
certain grid of time points.

Input to Stan (data):
- Time point grid (e.g. `sm_time_grid`)
- Length of the time point grid (e.g. `n_sm_time_grid`)

Output from Stan:
- Samples at observed time points (e.g. `log_surv_fit_at_obs_times`), this is a vector 
  in parallel to the existing `Times`. This can be in transformed parameters.
- Samples at time point grid (e.g. `log_surv_fit_at_time_grid`), this can be a matrix with
  `Nind` rows and `n_sm_time_grid` columns. This should happen in generated quantities
  because it is not needed for the log likelihood calculations and would slow
  us down otherwise unnecessarily. For now we would need to put that separately
  into each survival model's Stan code. 

We are doing this here on this branch to see how easy it is:
- In `DataSurvival.R` we need to 
    - add a `time_grid` slot and
    - importantly the `as.list()` method needs to return the required data points
- We need to modify `survival.stan` in the base folder
    - to arrange for the new inputs
    - to calculate the required survival fit samples

We note that the individual `model.stan` file in the corresponding sub-folder,
here for the log-logistic survival model, only contains the baseline hazard 
definition and the parameters. Therefore we don't need to change anything there.
This situation is actually easier here than for the longitudinal fit samples,
where the calculation is not yet centralized.

Finally we can get the samples:

```{r}
mcmc_results@results$summary("log_surv_fit_at_obs_times[1]")
mcmc_results@results$summary("log_surv_fit_at_time_grid[1,2]")
```

We definitely notice that the sampling speed is lower now with the additional
generated quantities generation.

### `survival` function as user interface

Here is the prototype.

```{r}
setGeneric(
    name = "survival", 
    def = function(object, ...) standardGeneric("survival")
)

.SurvivalSamples <- setClass(
    "SurvivalSamples",
    contains = "list"
)

setMethod(
    f = "survival",
    signature = c(object = "JointModelSamples"),
    definition = function(object, patients = NULL, ...) {
        all_pts <- names(object@data$pt_to_ind)
        if (is.null(patients)) patients <- all_pts
        assert_that(all(patients %in% all_pts), !any(duplicated(patients)))
        time_grid <- object@data$sm_time_grid
        assert_that(is.numeric(object@data$sm_time_grid))
        time_grid_index <- seq_along(time_grid)
        log_surv_fit_at_time_grid_samples <- object@results$draws("log_surv_fit_at_time_grid")
        log_surv_fit_at_obs_times_samples <- object@results$draws("log_surv_fit_at_obs_times")[, 1L, , drop = TRUE]
        results <- list()
        for (this_pt in patients) {
            this_result <- list()
            # Samples, also do exp() here.
            patient_ind <- object@data$pt_to_ind[this_pt]
            this_surv_fit_names <- paste0("log_surv_fit_at_time_grid[", patient_ind, ",", time_grid_index, "]")
            this_result$samples <- exp(log_surv_fit_at_time_grid_samples[, 1L, this_surv_fit_names, drop = TRUE])
            # Summary.
            surv_fit <- samples_median_ci(this_result$samples)
            this_result$summary <- cbind(time = time_grid, surv_fit)
            # Observations.
            this_t <- object@data$Times[patient_ind]
            this_death <- patient_ind %in% object@data$dead_ind_index
            this_surv_fit <- samples_median_ci(exp(log_surv_fit_at_obs_times_samples[, patient_ind, drop = FALSE]))
            this_result$observed <- data.frame(t = this_t, death = this_death, this_surv_fit)
            rownames(this_result$observed) <- this_pt
            # Save all.
            results[[this_pt]] <- this_result
        }
        .SurvivalSamples(results)
    }
)

# setMethod(
#     f = "[",
#     signature = "SurvivalSamples",
#     definition = function(x, i, ...) {
#         # Note that we cannot use `callNextMethod()` here because `list` is S3.
#         x@.Data <- x@.Data[i]
#         x
#     }
# )
```

Let's try it out:

```{r}
surv_fit_selected <- survival(mcmc_results, patients = c("pt_00011", "pt_00023")) 
surv_fit_all <- survival(mcmc_results)
# surv_fit_subset <- surv_fit_selected["pt_00011"]
```

This works quite ok, good! (again the subset does not work yet but let's see in the
package later...)

# Average in groups of patients (IDs)

Let's add an `aggregate` method now for `SurvivalSamples`.

- We just average the individual ID samples.
- Recalculate the summary.
- Row-bind the observations.

```{r}
setMethod(
    f = "aggregate",
    signature = c(x = "SurvivalSamples"),
    definition = function(x, groups, ...) {
        assert_that(
            is.list(groups), 
            !is.null(names(groups))
        )
        time_grid <- x[[1]]$summary$time
        results <- list()
        for (this_group in names(groups)) {
            this_result <- list()
            this_ids <- groups[[this_group]]
            # Samples.
            this_ids_samples <- Map("[[", x[this_ids], i = "samples")
            this_ids_samples <- lapply(this_ids_samples, "/", length(this_ids)) 
            this_result$samples <- Reduce(f = "+", x = this_ids_samples)
            # Summary.
            surv_fit <- samples_median_ci(this_result$samples)
            this_result$summary <- cbind(time = time_grid, surv_fit)
            # Observations.
            this_ids_obs <- Map("[[", x[this_ids], i = "observed")
            this_result$observed <- do.call(rbind, this_ids_obs)
            # Save all.
            results[[this_group]] <- this_result
        }
        .SurvivalSamples(results)
    }
)
```

Let's try it out:

```{r}
groups <- split(os_data$pt, os_data$arm)
surv_fit_grouped <- aggregate(surv_fit_all, groups = groups)
```

That seems to work nicely!

## `autoplot` method

Now let's define the plot method prototype on top of `SurvivalSamples`.

For now we are using the `ggkm` package to get a lightweight `geom_km` 
for adding a Kaplan Meier curve estimate to the plots. A disadvantage here is that
the package is not on CRAN yet. For production and before a CRAN release, we might 
want to influence the author Michael Sachs to also publish his package on CRAN
first.

The first part here is to stack together the `data.frame` parts in a long format
such that we can use `ggplot2` faceting.

```{r}
library(ggplot2)

# remotes::install_github("sachsmc/ggkm")

new_plot <- function(object, add_km = TRUE, ...) {
    assert_that(is.flag(add_km))
    
    all_fit_dfs <- lapply(object, "[[", i = "summary")
    all_fit_dfs_with_id <- Map(cbind, all_fit_dfs, id = names(object))
    all_fit_df <- do.call(rbind, all_fit_dfs_with_id)
    
    obs_dfs <- lapply(object, "[[", i = "observed")
    obs_dfs_with_id <- Map(cbind, obs_dfs, id = names(object))
    all_obs_df <- do.call(rbind, obs_dfs_with_id)
    # Due to https://github.com/sachsmc/ggkm/issues/5:
    all_obs_df$death_num <- as.numeric(all_obs_df$death) 
    
    p <- ggplot() +
        geom_line(aes(x = time, y = median), data = all_fit_df) +
        geom_ribbon(aes(x = time, ymin = lower, ymax = upper), data = all_fit_df, alpha = 0.3) +
        xlab(expression(t)) +
        ylab(expression(S(t))) +
        facet_grid(~ id)
    if (add_km) {
        p <- p + 
            ggkm::geom_km(aes(time = t, status = death_num), data = all_obs_df) +
            ggkm::geom_kmticks(aes(time = t, status = death_num), data = all_obs_df)
    }
    p
}

setMethod(
    f = "autoplot",
    signature = c(object = "SurvivalSamples"),
    definition = new_plot
)
```

Let's try it out:

```{r}
new_plot(surv_fit_grouped)
new_plot(surv_fit_grouped, add_km = FALSE)
```

We can also look at individual patients:

```{r}
new_plot(surv_fit_selected)
new_plot(surv_fit_selected, add_km = FALSE)
```

There is a warning from `ggplot2`, this seems from underlying code, see
https://github.com/sachsmc/ggkm/issues/11 

